\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-21 tex/paper.tex
add inline comments in intro + small changes

The discovery of new particles or physical phenomena in high energy
physics experiments, like those at the Large Hadron Collider, 
requires the observation of statistically significant deviations from
the predictions of the Standard Model, the current model
of the known fundamental particles and their interactions.  Typically,
discovery requires satisfying the \textit{5-sigma rule}, 
whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$.
The evaluation of the statistical significance must not only account
for statistical uncertainties, but also systematic 
uncertainties that represent our incomplete knowledge of the theory of
particle interactions or of the properties of the experimental
detection apparatus. Systematic uncertainties can alter the expected
rate of signal and background classes, e.g. their priors, as well as
the distributions of features. As such, the signal class true-positive
rate and the background class false-positive rate of a classifier can
change with variations in the systematic uncertainty thereby
decreasing the sensitivity of an analysis.

In this paper, we present a new strategy for training a classifier,
based on adversarial training~\citep{goodfellow2014generative}, 
which aims to build a classifier whose decision function is invariant
under variations due to systematic uncertainties.  In this way,
uncertainties on predictions from simulations will have a reduced
impact when computing the significance of potential deviations after
application of the discriminant model and thereby increase
the sensitivity of the analysis to signals from potential new physics sources.

Given training data $x \in X$ and associated labels $y \in \{s,
b_{i=1...k} \}$ where $s$ is the signal class and $b_{i=1...k}$ are
the background classes, then the probability distribution of the data
can be written
\[
p(x) = \pi_s p(x|y=s) + \sum_{i=1}^{k} \pi_{b_i} p(x | y=b_i)
\]
where $\pi_{y}$ represents the prior for the signal or a given
background, and $p(x|y)$ is the conditional distribution of the data
for the signal or a given background. Systematic uncertainties can
affect this distribution in the following ways:
\begin{itemize}
\item Uncertainties on the priors $\pi_{y}$.  These arise from
  uncertainties on the rates we expect a given process to occur as
  predicted from theoretical calculations or from alternative
  measurements of the rate of a given class.

\item Uncertainties in the simulators of the physical interactions,
  giving rise to uncertainties on the distributions $p(x|y)$.  These
  uncertainty are often determined by comparing the predictions of
  different simulations using differing models of the same physical
  process.

\item Uncertainties on the detection apparatus and it's effect on the
  measurement of particle properties, giving rise to uncertainties on
  the distributions $p(x|y)$.  For instance this includes
  uncertainties on our knowledge of the mean and variance of the
  measured energy distribution of a particle.

\end{itemize}
In this paper, we will not address uncertainties of the first type
mentioned above (uncertainties on the priors).  Rather, we will focus
on mitigating the effect of systematic uncertainties that alter the
conditional distributions $p(x|y)$.

Systematic uncertainties are parameterized by nuisance parameters in
the statistical analysis of data.  Nuisance parameter often, though
not always, have a suitable prior coming from alternative measurements
in data.  For instance, The average measured energy of an electron in
simulation may be compared with those in data using a relatively pure
sample of $Z$ boson decays to electrons.  The average measured
electron energy can not be measured without uncertainty, and the
measured uncertainty can be used in other analyses to estimate the
uncertainty deriving from our imperfect knowledge of the electron's
energy.  A nuisance parameter is assigned to the electron energy
uncertainty, which controls the variation of the electron energy in
the statistical analysis of data.  Nuisance parameters are typically,
though not always, constrained by a suitable prior, e.g. a Gaussian
distribution with a standard deviation equal to the measured uncertainty.

\glnote{Distinction between statistical and systematic uncertainty.}
\glnote{Outline taxonomy of uncertainties.}
\glnote{Define nuisance parameters. See refs in Noel's papers.}
\glnote{We want to build an accurate classifier whose output remains invariant with
respect to systematic uncertainties.}
% https://www.slac.stanford.edu/econf/C030908/papers/TUAT004.pdf
\glnote{Motivate the criterion (which may not be obvious for the ML crowd). See pivotal quantity motivation.}
\glnote{Discuss what kind of uncertainties the proposed approach is good for.}

>>>>

The discovery of new particles or physical phenomena in high energy physics
experiments, like those at the Large Hadron Collider, requires the observation
of statistically significant deviations from the predictions of the Standard
Model, the current model of the known fundamental particles and their
interactions.  Typically, discovery requires satisfying the \textit{5-sigma
rule}, whereby the deviation must be at least five standard deviations from
predictions, i.e. having a p-value of $p \lesssim 3 \times 10^{-7}$. The
evaluation of the statistical significance must not only account for statistical
uncertainties, but also systematic uncertainties that represent our incomplete
knowledge of the theory of particle interactions or of the properties of the
experimental detection apparatus. Systematic uncertainties can alter the
expected rate of signal and background classes, e.g. their priors, as well as
the distributions of features. As such, the signal class true-positive rate and
the background class false-positive rate of a classifier can change with
variations in the systematic uncertainty thereby decreasing the sensitivity of
an analysis. In this paper, we present a new strategy for training a classifier,
based on adversarial training~\citep{goodfellow2014generative}, which aims to
build a classifier whose decision function is invariant under variations due to
systematic uncertainties.  In this way, uncertainties on predictions from
simulations will have a reduced impact when computing the significance of
potential deviations after application of the discriminant model and thereby
increase the sensitivity of the analysis to signals from potential new physics
sources. \glnote{Shall we be more explicit somewhere and put a formula
underlining the effect of having a pivotal classifier on the resulting
sensitivity? (versus a classifier whose decision function depends on the
nuisance.)}

Given data $X$ and associated labels $Y$
taking values $y \in {\cal Y} = \{s,
b_{i=1...k} \}$ where $s$ is the signal class and $b_{i=1...k}$ are
the background classes, then the probability density of the data
can be written as a mixture
\begin{equation}
p(X) = \pi_s p(X|Y=s) + \sum_{i=1}^{k} \pi_{b_i} p(X | Y=b_i),
\end{equation}
where $\pi_{y}$ represents the prior for the signal or a given
background, and $p(X|Y)$ is the conditional distribution of the data
for the signal or a given background. Systematic uncertainties can
affect this distribution in the following ways \glnote{add references}:
\begin{itemize}
\item Uncertainties on the priors $\pi_{y}$.  These arise from
  uncertainties on the rates we expect a given process to occur as
  predicted from theoretical calculations or from alternative
  measurements of the rate of a given class.

\item Uncertainties in the simulators of the physical interactions,
  giving rise to uncertainties on the distributions $p(X|Y)$.  These
  uncertainties are often determined by comparing the predictions of
  different simulations using differing models of the same physical
  process.

\item Uncertainties on the detection apparatus and its effect on the
  measurement of particle properties, giving rise to uncertainties on
  the distributions $p(X|Y)$.  For instance, this includes
  uncertainties on our knowledge of the mean and variance of the
  measured energy distribution of a particle.
\end{itemize}
In this paper, we will not address uncertainties of the first type
mentioned above (uncertainties on the priors).  Rather, we will focus
on mitigating the effect of systematic uncertainties that alter the
conditional distributions $p(X|Y)$.

Systematic uncertainties are parameterized by nuisance parameters in the
statistical analysis of data \glnote{add references}.  Nuisance parameter often,
though not always, have a suitable prior coming from alternative measurements in
data.  For instance, The average measured energy of an electron in simulation
may be compared with those in data using a relatively pure sample of $Z$ boson
decays to electrons.  The average measured electron energy can not be measured
without uncertainty, and the measured uncertainty can be used in other analyses
to estimate the uncertainty deriving from our imperfect knowledge of the
electron's energy.  A nuisance parameter is assigned to the electron energy
uncertainty, which controls the variation of the electron energy in the
statistical analysis of data.  Nuisance parameters are typically, though not
always, constrained by a suitable prior, e.g. a Gaussian distribution with a
standard deviation equal to the measured uncertainty. \glnote{... and
accordingly, we want to build a classifier independent on the nuisance
parameter, as a way to be independent of the corresponding systematics.}

\glnote{Add paper outline.}
